---
title: "auk Full Workflow: AOS 2026 workshop"
author: "Aimee M. Van Tatenhove & Fabiola Rodríguez Vásquez"
format: html
fig-format: svg
editor: visual
theme: "flatly"
page-layout: full
toc: true
toc-location: left
embed-resources: true
execute:
  warning: false
  message: false
  out-width: "100%"
---

The full workflow for the AOS 2026 workshop, "Setting Up Your Analyses Up for Success with 'auk'". The workflow is divided into four lessons (i.e. separate scripts or Quarto documents) to:

1.  Learn how to import 'auk' data & general exploration of the files' structure.

2.  Learn how to use 'auk' filters to sample the eBird data for a project.

3.  Learn how to process the ***ebd*** and ***sampling*** files to obtain a dataset for species "presences-absences".

## Datasets and resources

Participants will find all relevant resources in the provided folder that includes an R project, data and codes.

### Data

The instructors downloaded and processed eBird data to prepare ebd and sampling files for the workshop.

- ***ebd*** and ***sampling*** files for three species found in the eastern United States, all available years.

  - ebd_filtered_us_AOS_2026.txt

  - effort_filtered_us_AOS_2026.txt

- ***ebd*** and ***sampling*** files for two species found in the seven Central American countries.

  - ebd_filtered_cam_AOS_2026.txt

  - effort_filtered_cam_AOS_2026.txt

Geospatial files will be used for the applications sections of this workshop and can be found in the folder "gis".

### Scripts

Separate Quarto documents (.qmd) or scripts? have been prepared for each of the objectives mentioned above. These documents have brief explanations or text that introduces an exercise and are followed by 'code chunks', spaces where R will be executed. Feel free to modify text below or above the code chunks to add your own notes, or add comments to your code within the code chunks.

## Import and explore eBird raw data

The first step in our workflow is to call the libraries needed. The libraries should be installed but in case you need to do so, apply install.packages('packages')

```{r}
library(tidyverse)
library(auk)
source("code/custom-functions.R")

theme_set(custom.theme())
```

### Section 1: Load EBD (eBird Basic Dataset) and effort (sampling) files

'auk' requires direction to where the downloaded eBird Basic Dataset (ebd) and effort (sampling) files are in your computer. Which is why this workflow needs to specify the directory of these input files.

```{r}
## Set paths for EBD & effort data in
in_ebd_small <- "./data/ebd_filtered_cam_AOS_2026.txt"
in_effort_small <- "./data/effort_filtered_cam_AOS_2026.txt"

```

The data has NOT been loaded yet! Remember that 'auk' is meant to be an effective way of processing the eBird large datasets but as we can see below these first sample datasets are not too big.

```{r}
## Smallish files, so we can load into R memory directly
get.filesize(in_ebd_small)
get.filesize(in_effort_small)

```

A function to read these files are *read_abd()* and *read_sampling().* Take a look at the help file for *read_ebd()*. It manages two things that are important to know: 1) The argument *unique=TRUE* , removes duplicated grouped checklists. 2) *rollup=TRUE,* deals with a taxonomic "rollup" so you are only working with species-level data.

```{r}
## How do we read the files?
?read_ebd()
```

Once the files are read in, we are ready to get familiar with these files. Notice that the ebd file is smaller than the effort file, why could that be?

```{r}
## Apply read_ebd()
ebd <- read_ebd(in_ebd_small) 
effort <- read_sampling(in_effort_small) 
```

### Section 2: Exploring the structure of the eBird data

The *ebd* object is the data for the species observed, *effort* holds the data for the sampling information. Compare their attributes (i.e. columns), to determine which are shared and which are not. 

```{r}
glimpse(ebd)
```

```{r}
glimpse(effort)
```

Notice that *ebd* has fewer rows than *effort*. This is because *ebd* comes from the file with species observations, *effort* does not have any species column, the data that was downloaded was all available checklist sampling information that matched the other criteria (location, dates). *effort* will be key to understand the prevalence of a species, given how many checklists were compiled for an area.

But before that lets keep exploring the *ebd* file because this is the data we are sampling from for our objectives. 

*How many species are available in this dataset?*
```{r}
unique(ebd$common_name)
```

*How many species?*
```{r}

```

*What is the date range for this data?*
```{r}
range(ebd$observation_date)
```

*How many counts does each species have?*
```{r}
table(ebd$common_name)
```

Now for some visual exploration. *Which country has the most detections?*
```{r}
## Explore visually by using ggplot's geom_bar which counts occurrences for a category
## Observations dataset
ggplot(ebd) +
  geom_bar(aes(x = country)) +
  ggtitle("Observations by country") +
  xlab("Country") + ylab("Count")
```

```{r}
ggplot(ebd) +
  geom_bar(aes(x = country, color = common_name, fill = common_name),position=position_dodge()) +
  ggtitle("Observations by Country and Species") +
  coord_flip()+
  xlab("Country") + ylab("Count") +
  labs(color = "Species",
       fill = "Species") 
```
*When does it seem to be a good time to observe the species?*
```{r}
ggplot(ebd) +
  geom_bar(aes(x = observation_date, color = common_name, fill = common_name)) +
  ggtitle("Observation Date by Species") +
  xlab("Checklist date") + ylab("Count") +
  labs(color = "Species",
       fill = "Species")
```
*Which types of observation types or counting protocols give more data?*
```{r}
## Effort dataset
ggplot(effort) +
  geom_bar(aes(x = observation_type)) +
  ggtitle("Observation Type") +
  coord_flip()+
  xlab("Type") + ylab("Count")
```

```{r}
ggplot(effort) + # You can safely ignore the plotting error here
  geom_bar(aes(x = duration_minutes)) +
  ggtitle("Checklist Duration") +
  xlab("Duration in minutes") + ylab("Count")
```

### Section 3: Pipes- the building block of the 'auk' workflow 
Pipes come in two flavors: base "|>" and magrittr (tidyverse) "%>%". We will use base pipes today.

```{r}
# Example: no pipes (stepwise)
tmp1 <- subset(effort, !is.na(duration_minutes))
tmp1 <- transform(tmp1, duration_hours = duration_minutes / 60)
duration_hours <- tmp1$duration_hours
duration_max_no_pipe <- max(duration_hours)

duration_max_no_pipe

# Example: with base R pipe 
duration_max_pipe <- effort |>
  subset(!is.na(duration_minutes)) |>
  transform(duration_hours = duration_minutes / 60) |>
  _$duration_hours |> # "_" passes object to next line of code
  max()

duration_max_pipe
```



## Setup and execute filters for eBird raw data in 'auk'

For this portion of the workshop we will use only the *ebd* file with species observations, to learn how 'auk' filters work.

```{r}
library(tidyverse)
library(auk)
source("code/custom-functions.R")
```

### Section 1: Set path

Remember, the first step is to setup the path for 'auk' to find the eBird files (ebd and sampling). We will continue working with the small Central American dataset.

```{r}
in_ebd <- "data/ebd_filtered_cam_AOS_2026.txt"
```

### Section 2: Define 'auk' filters

Defining 'auk' filters has been compared to conducting your sampling design with already collected data. **This** is where you customize the information you need from a large dataset such as a complete country's eBird dataset.

The package provides a prefix *'auk_'* to all functions that conduct a filtering task. 

If you are working with large datasets, the filtering step uses *auk_ebd* first, to reference the file of interest instead of reading it into your R session's memory. By building your entire filter pipeline you are effectively using AWK to help filter large files before having to load them in with *read_ebd*.

```{r}
?auk_ebd() 

# Try others such as ?auk_species, ?auk_date to see what they do
```

Lets build a simple set of filters. Remember that to build the sequence we will use the pipe "|>"

The output is a summary of your filters, in this case it indicates the species of interest as the only filter and the rest maintain "all" data.
```{r}
filters1 <-
  # Set EBD file path
  auk_ebd(in_ebd) |>
  # Species: common and scientific names can be mixed
  auk_species(species = "Crested Guan") 

filters1
```

As filters get more complicated (i.e. more than one species), it may be useful to create objects for the filters.
```{r}
# Define filters outside of pipeline; useful for complicated filters
species <- c("Crested Guan", "Northern Emerald-Toucanet")
species_names <- ebird_species(species, type = "common")
species_names

filters2 <-
  # Set EBD file path
  auk_ebd(in_ebd) |>
  # Species: common and scientific names can be mixed
  auk_species(species = species_names) 

filters2
```

Lets build more filters.
```{r}
## Exploring tailored filters: Thinking about sampling design and reducing variation
filters3<-auk_ebd(in_ebd) |>
  auk_species(species = species_names) |>
  auk_country("Honduras")|>
  # Date: use standard ISO date format `"YYYY-MM-DD"`
  auk_date(date = c("*-01-01", "*-12-31")) 

filters3
```

### Section 3: Apply filters and save new dataset

None of the filters above were saved in a file that can be used in analyses. The final step is to apply *auk_filter*.

Below we filter, save the data and load it with *read_ebd*
```{r}
# Determine the path and name of your output file
out_ebd<-"data/ebd_filtered_cam_presence.txt"

# Apply the filter
presence_out <- auk_filter(filters3,
                           file = "data/ebd_filtered_cam_presence.txt",
                           overwrite = TRUE) |>
  # Read filtered data into R environment
  read_ebd()
```

Below we confirm that this dataset only has the species and country data we specified.
```{r}
unique(presence_out$common_name)
unique(presence_out$country)
```
We can also check how much we have reduced an original eBird file when filtering to the needs. 
```{r}
get.filesize(in_ebd) 
get.filesize(out_ebd)
```



## Zerofilling eBird raw data to obtain a "presence-absence" dataset

Presence-only data are useful for exploring how frequently a species has been detected and where. However, for some analyses, they are not enough to understand what drives species occurrence when there is no information on where the species is absent. This is where **zerofilling** comes in, an 'auk' function that can take the eBird data file (*ebd*) and the sampling file (also called *sed* sometimes) to use the checklist_id information and fill in which of the sampling events a species was observed or not observed.

```{r}
library(tidyverse)
library(auk)
library(lubridate)
library(cowplot)
library(sf)
source("code/custom-functions.R")

theme_set(custom.theme())
```

### Section 1: Set the path for the files

For this zerofilling exercise we will use a different and larger dataset. In honor of the AOS 2026 location, we will use eBird raw data for the three banner species: American Woodcock, Prairie Warbler and Black-capped Chickadee. These observations were still filtered only to the eastern U.S. states and include all years of observation available.

```{r}
in_ebd <- "data/ebd_filtered_us_AOS_2026.txt"
in_effort <- "data/effort_filtered_us_AOS_2026.txt"

## Observe how large these files are compared to those from the previous section
get.filesize(in_ebd) 
get.filesize(in_effort)
```
### Section 2: Applying filters

After identifying the file paths, lets set up some filters. Note that when calling *auk_ebd*, we need to include the path for the sampling file.

**And**, we will also include a new function to ensure **complete** checklists using *auk_complete*. Complete checklists help confirm that we are capturing the most thorough sampling by an observer, including all available and identifiable species.
```{r}
species_names=c("amewoo")

filters <-
  # Set EBD & effort file paths
  auk_ebd(in_ebd, file_sampling = in_effort) |>
  # Species: common and scientific names can be mixed
  auk_species(species = species_names) |>
  # State: two-part 4-6 character codes
  ## (2-letter ISO country code & 1-3 character state code). E.g., "US-NY"
  auk_state(state = "US-MA") |>
  # Date: use standard ISO date format `"YYYY-MM-DD"`
  auk_date(date = c("2025-01-01", "2025-12-31")) |>
  # Time: 24h format with beginning and end times
  auk_time(start_time = c("17:00", "23:00")) |>
  # Complete: all species seen or heard are recorded (important!)
  auk_complete()
filters
```

We filter the eBird and sampling data, save it and upload these new (*filtered*) data files.

```{r}
f_ebd <- "data/ebd_filtered_amewoo_presabs.txt"
f_effort <- "data/effort_filtered_amewoo_presabs.txt"

## Call AWK to execute filters
f_ebd_effort <- auk_filter(filters,
                          file = f_ebd,
                          file_sampling = f_effort,
                          overwrite = TRUE) |>
  read_ebd()

glimpse(f_ebd_effort) 
```

### Section 3: Zerofill to generate the presence-absence dataset

Execute the zerofill function *auk_zerofill*.

```{r}
presabs_zf <-auk_zerofill(f_ebd,
              sampling_events=f_effort,
               collapse = TRUE)

glimpse(presabs_zf)
```

The columns *observation_count* includes counts for each sampling event (i.e. *sampling_event_identifier* column).
```{r}
glimpse(presabs_zf$observation_count) #observation
glimpse(presabs_zf$sampling_event_identifier) #for each sampling event
```
While this is already the presence-absence dataset there is some processing necessary for the columns. For example take a look at the range of the counts:
```{r}
range(presabs_zf$observation_count)
```
"X" is a valid recording character used by observers when they: do not want to count all individuals or are unsure of how many indivivuals were detected. Because there is no way of guessing if X means 1 or more, we should remove these from future analyses. 

There are other small tweaks needed on the data for better comprehension and use for future analysis.

```{r}
zf_eff_transf <- presabs_zf |>
  mutate(
    # Convert count to integer and X to NA (ignore NA warning!)
    ## X indicates species was present, but a number wasn't recorded
    observation_count = as.integer(observation_count),
    # effort_distance_km to 0 for stationary counts
    effort_distance_km = if_else(observation_type == "Stationary",
                                 0, effort_distance_km),
    # Convert duration to hours
    effort_hours = duration_minutes / 60,
    # Convert speed km/h
    effort_speed_kmph = effort_distance_km / effort_hours,
    # Split date into year and day of year
    month = month(observation_date),
    day_of_year = yday(observation_date)
  )
```

Along the way in the filtering process we can take a look at the distribution of these variables to learn about how much data we have.

*In how many checklists was our species seen?*
```{r}
table(zf_eff_transf$species_observed)
```

*What is the range of the effort in distance covered?*
```{r}
ggplot(zf_eff_transf) +
  geom_bar(aes(x = effort_distance_km)) +
  ggtitle("Sampling effort: distance covered") +
  xlab("Distance (km)")
```
*What is the range of the effort of hours of observation?*
```{r}
ggplot(zf_eff_transf) +
  geom_bar(aes(x = effort_hours)) +
  ggtitle("Sampling effort: hours of observation") +
  xlab("No. hours")
```
*For which time periods is there data?*
```{r}
ggplot(zf_eff_transf) +
  geom_bar(aes(x = observation_date)) +
  ggtitle("Sampling effort: month") +
  xlab("Month")
```

These graphs will help determine whether there are time periods, types of protocols (stationary versus walked), or effort levels that should be filtered out because they represent low data amounts.

```{r}
zf_eff_filtered <- zf_eff_transf |>
  filter(observation_type %in% c("Stationary", "Traveling"),
         !is.na(effort_hours), effort_hours >= 0.17, effort_hours <= 5,
         !is.na(effort_distance_km), effort_distance_km <= 10,
         effort_speed_kmph <= 50,
         number_observers <= 10)

# Compare the filtered and transformed datasets
table(zf_eff_transf$species_observed)
table(zf_eff_filtered$species_observed)
```
While the prevalence of species did not increase too much, a visual below shows how the data is better constrained to an effort window.
```{r}
plot_grid(
  ggplot(zf_eff_transf) +
    aes(x = effort_hours) +
    geom_histogram(binwidth = 0.5,
                   aes(y = after_stat(count / sum(count)))) +
    scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
    labs(x = "Duration (hours)",
         y = "% of eBird checklists",
         title = "Distribution of eBird checklist duration",
         subtitle = "Before effort filtering"),
  
  ggplot(zf_eff_filtered) +
    aes(x = effort_hours) +
    geom_histogram(binwidth = 0.5,
                   aes(y = after_stat(count / sum(count)))) +
    scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
    labs(x = "Duration (hours)",
         y = "% of eBird checklists",
         title = "Distribution of eBird checklist duration",
         subtitle = "After effort filtering"),
  ncol = 2)
```

This zerofilled data is almost ready for an analysis or further exploration but one last thing to consider after all the desired filters have been applied is to clean the data from columns that are not necessary.
```{r}
zf<-zf_eff_filtered|>
  select(checklist_id,
         latitude,
         longitude,
         observation_date,
         effort_distance_km,
         number_observers,
         scientific_name,
         observation_count,
         species_observed,
         effort_hours,
         month,
         day_of_year)
```


### Section 4: Final visualization
Because each checklist or sampling event has latitude/longitude data, this data can be presented in maps and in future analyses, it can be linked to georefferenced environmental covariates.

```{r}
# Spatialize zerofilled object
zf_sf <- zf |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Load GIS and shapefile data
ee_landcover <- read.csv("data/gis/EarthEnv-landcover-northeast.csv") |>
  left_join(landcover_palette) # Add map colors for nice plotting

usa_sf <- read_sf("data/gis/usa-all-states.shp") |>
  st_transform(crs = 4326)

ne_states_sf <- read_sf("data/gis/usa-northeast-states.shp") |>
  st_transform(crs = 4326)
```
ll
```{r}
ggplot() +
  geom_sf(data = ne_states_sf) +
  geom_sf(data = zf_sf, aes(color = species_observed), alpha = 0.5)

```
Lets modify the map to visualize the state of interest.
```{r}

state_sf<-ne_states_sf[ne_states_sf$NAME=="Massachusetts",]
# Clip checklists to land; safe to ignore warning message
zf_clip_sf <- st_intersection(zf_sf, state_sf)

## Clipped map
ggplot() +
  geom_sf(data = state_sf) +
  geom_sf(data = zf_clip_sf, aes(color = species_observed), alpha = 0.5) +
  coord_sf(xlim = c(-74, -69), ylim = c(41, 43)) +
  ggtitle("eBird Observations by Checklist") +
  labs(color = "Species observed")
```

With your dataset you can plot it alongside other spatial datasets of interest such as this example with landcover.
```{r}
ggplot() +
  geom_raster(data = ee_landcover,
              aes(x = x, y = y,
                  fill = class_names), alpha = 0.7) +
  geom_sf(data = ne_states_sf, color = "grey30", fill = NA) +
  geom_sf(data = zf_clip_sf, aes(color = species_observed), alpha = 0.5) +
  scale_fill_manual(breaks = unique(ee_landcover$class_names),
                    values = unique(ee_landcover$colors)) +
  labs(color = "Species observed",
       fill = "Landcover class")
```

