---
title: "03- Generating 'presence-absence' datasets"
author: "Aimee M. Van Tatenhove & Fabiola Rodríguez Vásquez"
format: html
editor: visual
---

## Goal of section 3

Learn how to process the eBird raw data (***ebd***) and checklists events (***sampling***) files to obtain a dataset for species "presences-absences".

Presence-only data are useful for exploring how frequently a species has been detected and where. However, for some analyses, they are not enough to understand what drives species occurrence when there is no information on where the species is absent. This is where **zerofilling** comes in, an 'auk' function that can take the eBird data file (*ebd*) and the *sampling* file to use the checklist_id information and fill in when a species was or not observed.

```{r Set section 3 R environment}
library(tidyverse)
library(auk)
library(lubridate)
library(cowplot)
library(sf)
source("custom-functions.R")

theme_set(custom.theme())
```

### 3.1. Set the path for the files

For this zerofilling exercise we will use a different and larger dataset. In honor of the AOS 2026 location, we will use eBird raw data for the three banner species: American Woodcock, Prairie Warbler and Black-capped Chickadee. Observations were filtered for the eastern U.S.

```{r Setting the path to a new dataset}
# Set the paths for the US AOS 2026 dataset:
in_ebd <- 
in_effort <- 

# Observe how large these files are compared to those from the previous section
get.filesize(in_ebd) 
get.filesize(in_effort)
```

### 3.2. Applying filters

After identifying the file paths, let's set up some filters. Note that when calling *auk_ebd*, we need to include the path for the sampling file.

**And**, we will also include a new function to only include **complete** checklists using *auk_complete*. Complete checklists help confirm that we are capturing the most thorough sampling by an observer, including all available and identifiable species.

```{r Setting filters new dataset}

# Prepare the species filtering object for the American Woodcock
species_names = c("amewoo")

filters <-
  # Set EBD & effort file paths:
  
  # Set the species object filter:
  
  # Set a filter for the US state of Massachusetts. E.g., "US-NY"
  
  # Set a filter for date range between January 1, 2025 and December 12, 2025:
  
  # Set a filter for time range between 5 and 11 pm (Time follows a 24h format with beginning and end times):
  
  # Always call auk_complete: all species seen or heard are recorded (important!)
  auk_complete()

filters
```

We filter the eBird and sampling data, save it and upload these new (*filtered*) data files.

```{r Apply filters new dataset}
# Set the path and filename for the ebd and effort output files:
f_ebd <- "../data/eBird/ebd_filtered_amewoo_presabs.txt"
f_effort <- "../data/eBird/effort_filtered_amewoo_presabs.txt"

  # Filter the data using the filtering object, indicate the output files: 
f_ebd_effort <- auk_filter() |>
  # Call read_ebd
  read_ebd()

glimpse(f_ebd_effort) 
```

### 3.3. Zerofill to generate the presence-absence dataset

Execute the zerofill function *auk_zerofill*.

```{r Apply zerofill}
# Call zerofill including the effort and ebd files:
presabs_zf <-auk_zerofill()

glimpse(presabs_zf)
```

The columns *observation_count* includes counts for each sampling event.

```{r Check columns with presence-absence}
glimpse(presabs_zf$observation_count) # observation
glimpse(presabs_zf$checklist_id) # for each sampling event
```

While this is already the presence-absence dataset there is some processing necessary for the columns to be useful for future analyses. For example take a look at the range of the counts:

```{r Exploring variable for counts}
range(presabs_zf$observation_count)
```

"X" is a valid recording character used by observers when they: do not want to count all individuals or are unsure of how many individuals were detected. Because there is no way of guessing if X means 1 or more, we should remove these for future analyses.

There are other small tweaks needed on the data for better comprehension and use for future analysis.

```{r Clean presence-absence dataset}
zf_eff_transf <- presabs_zf |>
  mutate(
    # Convert count to integer and X to NA (ignore NA warning!):
    observation_count = as.integer(observation_count),
    # Effort_distance_km to 0 for stationary counts:
    effort_distance_km = if_else(observation_type == "Stationary",
                                 0, effort_distance_km),
    # Convert duration to hours:
    effort_hours = duration_minutes / 60,
    # Convert speed km/h:
    effort_speed_kmph = effort_distance_km / effort_hours,
    # Split date into year and day of year:
    month = month(observation_date),
    day_of_year = yday(observation_date)
  )
```

Along the way in the filtering process we can take a look at the distribution of these variables to learn about how much data we have.

*In how many checklists was our species seen?*

```{r Explore balance of observations}
table(zf_eff_transf$species_observed)
```

*What is the range of the effort in distance covered?*

```{r Plot effort by distance}
ggplot(zf_eff_transf) +
  geom_bar(aes(x = effort_distance_km)) +
  ggtitle("Sampling effort: distance covered") +
  xlab("Distance (km)")
```

*What is the range of the effort of hours of observation?*

```{r Plot effort by hours}
ggplot(zf_eff_transf) +
  geom_bar(aes(x = effort_hours)) +
  ggtitle("Sampling effort: hours of observation") +
  xlab("No. hours")
```

*For which time periods is there data?*

```{r Plot effort by period}
ggplot(zf_eff_transf) +
  geom_bar(aes(x = observation_date)) +
  ggtitle("Sampling effort: month") +
  xlab("Month")
```

These graphs will help determine whether there are time periods, types of protocols (stationary versus walked), or effort levels that should be filtered out because they represent low data amounts.

```{r Tweaking filters}
zf_eff_filtered <- zf_eff_transf |>
  # Filter to include only stationary and traveling protocols:
  filter(observation_type %in% c("", ""),
         # To include between up to 5 hours:
         !is.na(effort_hours), effort_hours >= 0.17, effort_hours <= ,
         # To include distances up to 10 km:
         !is.na(effort_distance_km), effort_distance_km <= ,
         # To include up to 50 km/h:
         effort_speed_kmph <= ,
         # To include up to 10 observers:
         number_observers <= )

# Compare the filtered and transformed datasets
table(zf_eff_transf$species_observed)
table(zf_eff_filtered$species_observed)
```

While the prevalence of species did not increase too much, a visual below shows how the data is better constrained to an effort window.

```{r Checking dataset coverage}
plot_grid(
  ggplot(zf_eff_transf) +
    aes(x = effort_hours) +
    geom_histogram(binwidth = 0.5,
                   aes(y = after_stat(count / sum(count)))) +
    scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
    labs(x = "Duration (hours)",
         y = "% of eBird checklists",
         subtitle = "eBird checklists duration\nbefore effort filtering"),
  
  ggplot(zf_eff_filtered) +
    aes(x = effort_hours) +
    geom_histogram(binwidth = 0.5,
                   aes(y = after_stat(count / sum(count)))) +
    scale_y_continuous(limits = c(0, NA), labels = scales::label_percent()) +
    labs(x = "Duration (hours)",
         y = "% of eBird checklists",
         subtitle = "eBird checklists duration\nafter effort filtering"),
  ncol = 2)
```

This zerofilled data is almost ready for an analysis or further exploration but one last thing to consider after all the desired filters have been applied is to clean the data from columns that are not necessary.

```{r Final preparations for auk clean dataset}
zf <- zf_eff_filtered |>
  # Select only columns that you would use in your analysis:
  select(checklist_id,
         latitude,
         longitude,
         observation_date,
         effort_distance_km,
         number_observers,
         scientific_name,
         observation_count,
         species_observed,
         effort_hours,
         month,
         day_of_year)
```

### 3.4. Final visualization

Because each checklist or sampling event has latitude/longitude data, this data can be presented in maps. It can also be linked to georeferenced environmental covariates for future analyses.

```{r Prepare to visualize data spatially}
# Spatialize zerofilled object
zf_sf <- zf |>
  st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# Load GIS and shapefile data
ee_landcover <- read.csv("../data/gis/EarthEnv-landcover-northeast.csv") |>
  left_join(landcover_palette) # Add map colors for nice plotting

ne_states_sf <- read_sf("../data/gis/usa-northeast-states.shp") |>
  st_transform(crs = 4326)
```

Plot the locations where the species was observed.

```{r Plot distribution of observations}
ggplot() +
  geom_sf(data = ne_states_sf) +
  geom_sf(data = zf_sf, aes(color = species_observed), alpha = 0.5)

```

Lets modify the map to visualize the state of interest.

```{r Plot observations for site of interest}
state_sf <- ne_states_sf[ne_states_sf$NAME == "Massachusetts", ]
# Clip checklists to land; safe to ignore warning message
zf_clip_sf <- st_intersection(zf_sf, state_sf)

## Clipped map
ggplot() +
  geom_sf(data = state_sf) +
  geom_sf(data = zf_clip_sf, aes(color = species_observed), alpha = 0.5) +
  coord_sf(xlim = c(-74, -69), ylim = c(41, 43)) +
  ggtitle("eBird Observations by Checklist") +
  labs(color = "Species observed")
```

With your dataset you can plot it alongside other spatial datasets of interest such as this example with landcover.

```{r Plotting observations with other spatial objects}
ggplot() +
  geom_raster(data = ee_landcover,
              aes(x = x, y = y,
                  fill = class_names), alpha = 0.7) +
  geom_sf(data = ne_states_sf, color = "grey30", fill = NA) +
  geom_sf(data = zf_clip_sf, aes(color = species_observed), alpha = 0.5) +
  scale_fill_manual(breaks = unique(ee_landcover$class_names),
                    values = unique(ee_landcover$colors)) +
  xlab("Longitude") + ylab("Latitude") +
  labs(color = "Species observed",
       fill = "Landcover class")
```

Try zooming in (using *coord_sf*) to see if you can identify any associations with observations of your species and landcover types.

```{r Plotting observations with other spatial objects zoom}
ggplot() +
  geom_raster(data = ee_landcover,
              aes(x = x, y = y,
                  fill = class_names), alpha = 0.7) +
  geom_sf(data = ne_states_sf, color = "grey30", fill = NA) +
  geom_sf(data = zf_clip_sf, aes(color = species_observed), alpha = 0.5) +
  coord_sf(xlim = c(-73, -72), ylim = c(42, 42.5)) + # Adjust to your liking
  scale_fill_manual(breaks = unique(ee_landcover$class_names),
                    values = unique(ee_landcover$colors)) +
  xlab("Longitude") + ylab("Latitude") +
  labs(color = "Species observed",
       fill = "Landcover class")
```

End of section and workshop! **Explore the applications examples and best of luck in your future analyses with eBird data :)**
